draft 1

### 2016-07-27 finalwriteup/draft.rmd
abstract

* why did you do the study?, what is the problem
performing loss-making jobs for fixed fees is a major problem in the consulting, particularly in the competitive construction industry.

* why is it important?
affects the porfitability of a company which impacts the careers, internal morale, and even viability of a company.

* why should someone read your essay?
learn if a company's passively collected data could help them avoid loss making jobs or tactfully enforce strictier contracts.

* what is your main claim?
improve profitability by 9% by avoiding ~4% of projects is a possible way to interpret the agorithm output. other ways are possible.

* discuss variables and your approaches
variables: main employee, % time by each type of employee, team size, client charactersitcs, project characteristics (size, type of work, length)
approaches: complex data with mnay variables, thousands of data points. tried regression as well as machine learning algs: randomf roest and boosted trees, bayesian networks, naive bayes.

@@

* overview of most important sources

@@

As stated, the aim of this project is to use statistical and machine learning techniques to model the profitability of projects. Several techniques will be tested and this section reviews a range of predictive algorithms that have been successfully applied to business and social data. These vary from simple methods such as Linear Regression to complex, deep learning Neural Networks. The review of methods applied to the construction and IT industry highlighted the use of Linear Regression, Neural Networks, SVM's and in one case Boosted Trees for prediction, however research on other business problems utilised a wider range of methods. These include Naive Bayes, Random Forests, and machine learned Bayesian Networks. The following section presents each method's advantages and disadvantanges and an example of a successful application.

@@

Linear Regression is one of the simplest and most popular statistical prediction methods. An equation for a line is learned, which is defined by a constant and coefficients multiplied by each explanatory variable which are assumed to be independent. This can be mathematically represented by the equation below:

@@

where $X_{j}$ represents the jth input variable, $\beta_{j}$ is the coefficient or parameter being solved for, and $\beta_{0}$ is the y-intercept. The most popular method to find the coefficients is to minimise the residual sum of squares (ordinary least squared method). The residuals represent the Y distance of each point from the line (predicted values), and the residual sum of squares is the sum of the squared value of each point's residual. It can be represented by the equation below:

@@

The reason for this method's popularity is that it is simple, provides stable predictions, and can be adapted for categorical variables [@Seng2010]. The integration of categorical variables is called Analysis of Variance (ANOVA) Linear Regression, and the algorithm compares group mean variances within the categorical variables.

If structured correctly, linear regression results can show the importance of the explanatory variables. The variables must be standardised to one another (i.e. scaled to the same mean and standard deviation), and then the coefficients (the slopes) indicate how *much* a variable changes with the response variable. This relationship can then be compared against the other explanatory variables [@Putler2012]. However, if two or more predictor variables are correlated, the values of their coefficients are unstable. More specifically, this means that if the a model has two correlated variables, and is re-run with a slightly different set of data, the coefficients for the two variables may be quite different because the assumption of independent variables is not true. On the other hand, correlated variables do not impact the predictive accuracy of the model [@Putler2012].

In linear regression, no assumptions are made about the distributions of the explanatory variables or response variable. Skewed explanatory variable distributions are acceptable but it is generally better to normalise them by, for example, taking the logarithm or square root of the variable. The reason for this is that the more extreme values from the skewed distribution may create high leverage points and influence the slope (variable coeffiecients) too strongly [@Chatterjee1986]. The distribution of the residuals on the other hand, should always be Gaussian with a mean of 0. This is because if modeled well, the regression equation should follow the shape of the data [@Hastie2009].

Unfortunately many real world world phenomena do not have linear relationships which can make it difficult to produce accurate results using this method [@Breiman2001b]. There are methods to overcome this limitation such as including polynomial terms and interaction terms. For example, in the linear equation, as well as $X_{1}$, include the terms $X_{2} = X_1^2$ and $X_{3} = X_1^3$ to build a third degree polynomial. Interactions between explanatory variables can be modelled by including a term such as $X_5$ where $X_5 = X_4*X_1$. This versatility is advantageous, however it can be challenging to know which non-linear terms of which variables to include. Sometimes data cannot be closely modelled by any equation and this is where other non-linear non-parametric methods can outperform linear regression.

@@

The ordinary least squared regression method described in the previous section cannot be applied to binary classification problems, where the objective is to predict a probability of an event occurring. Probability is a continuous response variable, however predictions from a linear regression would not be bound by 0 and 1. Instead, predicted values could reach further along the line outputting probabilities greater than 1 and less than 0 which is difficult to interpret. Also, residuals from linear regression should be normally distributed, but this is not possible when the response variable consists of only two values (**refappstate**).

Nevertheless, the theory behind Linear Regression can be adapted to the prediction of probabilities pertaining to a binary response variable (1 or 0). First, a sigmoid function is assumed as the functional form of the probabilties against variable X. This is because it is pragmatic and works well: sigmoid functions can be bound by 0 and 1, and a logit transformation of the probability results in a linear relationship between the log odds and X. Refer to the equation below where *p* represents probability:

@@

The aim is to find which category has a higher probability, so to simplify calcualtions, the denominator can be dropped. Because it is the same for both equations, the denominator does not affect the final ratio of probability of 'yes' to probability of 'no'. Integrating the assumption of conditional independence, the equations can be written as follows:

@@

These equations are simple to compute given the data, and the class with the highest probability is chosen [@Provost2013].

The advantages of this method are that the conditional independence assumption enables very fast calculations and predictions. The method can perform well in real world tasks because the assumption of independence does not significantly damage predictions. If multiple variables are actually related, the variables separately still direct the prediction in the correct direction. The correlated variables will double or triple their emphasis on the predicted response variable, however this does not necessarily impact classification accuracy because the final step is to choose the class with the *highest* probability [@Provost2013]. This is fine for ranking but the output probabilities are not accurate statistical probabilities [@Caruana2006]. Another disadvantage is that the distribution of the numeric variable must be assumed (often Gaussian) in order to calculate the liklihood terms which the data may not fit into neatly [@Caruana2006].

The Gaussian probability densities of two response classes can be visualised on a bivariate feature space. The two normal distrubtions represent the probability of each class given values of the two explanatory variables (x and y axes):

@@

One real world success story for this classifier hinged on its ability to be refined incrementally with each piece of new data, as opposed to re-calibrating the entire model. Naive Bayes was used in a complex spam detection system where new spam emails or toxic text themes could be quickly added to the filtering model. Although the example is quite different to cost estimation, the Naive Bayes method generally provides a good benchmark to compare against more complex models that should outperform it [@Caruana2006].

@@

Decision trees are one of the simplest and most intuitive machine learning methods. There are several brands of basic decision tree algorithms including ID3, C4.5, CART, and CHAID with CART and C4.5 being the most popular [@Kabra2011]. Ravi Kumar and Ravi [@Kumar2007] recommended the CART algorithm as it is capable of solving both classification and regression problems whereas the remaining decision trees solve classification problems only. Single decision trees create intuitive rules that a decision maker can follow in real-life scenarios but on the other hand, they tend to overfit the data and provide low predictive accuracy [@Kabra2011][@Putler2012]. Ensemble trees were developed to solve this problem and several ensemble techniques are described in this section.

Decision trees have a different theoretical foundation to traditional prediction methods such as regression. They are created by a progression of binary splits of the dataset based on response variable values. Before a split is made, each explanatory variable and each value within the explanatory variables is tested for how 'purely' it partitions the response variable into two subsets. For example, the explanatory variable 'gender' might split a response variable 'hair length' of a class of students into two subsets of students with more similar hair lengths. This similarity is called 'purity' and can be measured using several formulas. An example for regression is:

$SSN - (SSL + SSR)$

@@

where SSN is the sum of squared errors of a node (dataset before the split). $y_i$ are the response variable values and $\bar{y}$ is the mean response variable value in the node. SSL and SSR are the sum of squared errors in the left and right datasets that resulted from the split. Other purity formulas exist for categorical response variables such as Gini purity and information gain [@Hastie2009]. After the first split is made on the dataset, each subset is split by another variable (or the same explanatory variable again) and so on resulting in final subsets of data with similar response variable values. If this process is visualised graphically, it is clear that non-linear relationships can be captured.

@@

Once the splitting rules have been learned, a new case is run through the tree from the top. It follows the path provided by which side of the binary split its explanatory variable values fall under and ends up in a final node. The mean response variable value from the initial data that ended up in that node is the predicted response value for the new case. Decision trees are easy to overfit on the training data. This means that if the splitting rules are continued until no more splits can be possibly made, when the test data is run through the tree, the predictions will be too closely aligned to the exact structure of the training dataset. Prediction rules need to generalise trends found in the training data. This can be managed in a single tree by using 'stopping rules' such as a minimum number of data in a final node or a minimum purity increase for a split to be valid (**ref**).

As stated previously, single decision trees also suffer from low predictive accuracy and instability, so to combat these problems, ensemble tree methods were pioneered in the 1990's [@Breiman1996]. Three examples of these methods are boostrap aggregating (bagging), Random Forests, and boosting. Bagging builds multiple trees from different training sample data sets (wich are sampled from the full data set *with* replacement) and then combines tree results [@Breiman1996]. The method of combination is generally averaging numeric predictions or taking the majority rule for categorical variables.

@@

The boosted decision tree approach applies gradient descent theory to a series of decision trees. The trees are limited to a certain depth to maintain simplicity, and each tree models the residuals (or errors) of the preceding tree. By modeling the errors, misclassified cases are weighted higher than correctly classified cases and influence the structure of the tree more [@Hastie2009]. This increased weighting on errors is why the algorithm is called boosting.

This idea can be visualised in a simple scenario consisting of a single explanatory variable (shown on x-axis) and response variable (y-axis). Each tree is limited to one split, and the first tree is shown below, where the vertical line is the splitting point:

@@

The horizontal lines indicate the mean response variable value in each subset after the split. It is the value the first tree would predict for data that falls into each subset. The next tree is built on the residuals of the preceding tree.

![Second tree modeling residuals in short progression of boosted trees [@Ihler2012].](images/lit_review/boost_tree2_.png)

@@

The advantages of ensemble tree methods are numerous. Similarly to Linear Regression, the predictor variables in trees do not need to be transformed as no assumptions need to be made about the data's statistical distributions [@Louppe2014][@Radenkovic2010]. However, it is often wise to transform skewed distributions to achieve more even segmentation. Additionally, ensemble tree methods are faster than SVM's and Neural Networks but perform just as well and even provide insights such as variable importance and variable relationships [@Sealfon2012]. Random forests in particular have the benefit of being unaffected by noise in the data and has the capacity to handle a large number of variables [@Sealfon2012]. Caruana [@Caruana2006] tested Boosted Trees, Random Forests, Neural Networks, SVM's, Logistic Regression and Naive Bayes on 11 binary classification problems and found that Boosted Trees performed best, followed by Random Forests. This demonstrates how ensemble tree methods are capable of competing with high level machine learning algorithms.

Although ensemble trees have numerous advantages it is just as important to understand their limitations. Trees are not built on a probabilistic framework, and therefore their results can not be provided in this framework. For example, statistical confidence intervals for predictions are not available for standard ensemble methods [@Louppe2014]. This will be addressed again in the summary for this section. Also, variable importance tables provided by Random Forests and Boosted Trees can be biased towards variables with many categories and another type of ensemble method, conditional forests, should be used [@Radenkovic2010]. The methods can also be prone to overfitting if not tuned carefully [@Louppe2014][@Caruana2006].

@@

A Bayesian Network is a graphical probabilistic model that illustrates the conditional dependencies between variables in a data set. The model is visually represented by a directed acyclic graph (DAG) and is capable of linking the conditional dependency between any variable to another variable. The links are graphically represented by arcs or edges [@Heckerman1998]. This kind of conditional relationship is Bayesian, where the probabilities in one node are conditional upon values in nodes directed to it as well as preceding nodes. Refer to the simple example below:

@@

If each variable, A, B, C, and D are binary categories, there are $2^4$, or 16 combinations of results and each variable is linked to each of the other variables with a conditionally dependency. The probability of each combination could be calculated as per below using the chain rule for probabilities:

@@

However, if it is known through domain knowledge or can be reasoned through the data that some variables are conditionally independent from one another, the calculations can be simplified [@Barber2012]. For example, if B and D, A and D, and A and C are conditionally independent, the DAG would be:

@@

In this way, Bayesian Networks reduce computations required to find the probability of an event given other explanatory variable values while still taking into account many conditional dependencies that Naive Bayes cannot [@Barber2012]. Calculating the conditional probability tables for each variable is still much more computationally expensive than Naive Bayes though (as Naive Bayes is the simplest form of a Bayesian Netwoork) and variables with many categories dramatically increases the expense [@Zhang2004].

@@

Drawbacks of Bayesian Networks include the difficulty for them to process continuous variables. These demand sampling from conditional density models or complex algebra whereas categorical variables require comparitively simple calculations [@Hofmann1996]. It is therefore common to discretise continuous variables [@Kragt2009].

Bayesian networks have found success in combining deterministic models with observed data as well as expert knowledge. They are excellent at combining knowledge from different origins, and then visually communicating results well to decision makers [@Kragt2009]. Therefore Bayesian Networks have become an excellent decision support tool. Another benefit of Bayesian Networks as decision support tools is they allow the user to fix certain variable values, re-run the network and observe changes in probabilities. Thisis a kind of sensitivity testing that gives users a feel for whether the model makes sense and the degree of uncertainty. Popular applications of Bayesian Networks are modelling uncertainty in natural resource management and modelling complex business network structures such as airports [@Wu2013].

@@

Algorithmic Neural Networks were inspired by Neural Networks of the human nervous system. Biological Neurons recieve electro-chemical input from the terminals of adjacent neurons into its dendrites. If enough electro-chemical input is recieved, the neuron will fire, which means a signal will be sent from the dendrite end, through the axon to its terminal. Its terminal then sends signals to adjacent dendrites of other neurons and the process propagates. Neurons are able to learn over time by strengthening the connections of adjacent neurons that frequently cause the next one to fire. This means the required firing threshold for the recieving neuron lowers [@Putler2012].

@@

In fact, a wide variety of neural network structures and dynamics exist so a simple popular version is explained here. The first input layer is passive, and each input variable is passed through one input node. These nodes are passive and the variable values are simply passed to each of the nodes in the next layer, called a hidden layer. Neural networks may have any number of hidden layers with any number of nodes as selected by the analyst. Often however, a single hidden layer is adequate. Calibration for predictions is done in the hidden nodes. At these nodes, each input variable is weighted then combined (similar to linear regression) to produce a single value. This is then transformed and assessed by a threshold function to determine if the hidden node should 'fire' to the next nodes. Once the signals and values have propogated to the final layer they are recombined to produce an output, or a prediction [@Putler2012].

The algorithm learns from the data by starting with trial weights for each node. This produces a set of predictions from which errors are calculated. Then optimisation techniques are used to adjust the weights and reduce the error. This process can result in overfitting, so stopping rules such as regularisation strength should be used [@Karpathy2016].

@@

SVM's are non-probabalistic learning algorithms that are typically used for binary classification, although multi-class and regression problems can be solved. It works by geometrically determining the best decision boundary, called a hyperplane, between two classes in multi-dimensional space [@Byun2002] . A hyperplane is simply a plane in more than 3 dimensions, i.e. more than 3 explanatory variables. The simplest type of SVM resembles Logistic Regression because the hyperplane is a linear boundary, and like Logistic Regression, the category of future cases depends on which side of the boundary they fall.

@@

The plane is expressed with vectors because they can be defined such that $\boldsymbol{w}$ is perpendicular to $\boldsymbol{x}$. This aids in optimisation calculations to find the hyperplane with the widest margin, i.e. two hyperplanes with the largest perpendicular distance between them and less than the allowable misclassified points [@Fletcher2009]. The plot below demonstrates this concept and has no misclassifications:

@@

This section reviewed the advantages and disadvantages of statistical and machine learning techniques that have been successfully applied to business problems. These included Linear Regression, Logistic Regression, Naive Bayes, decision trees and ensemble trees, Bayesian Networks, Neural Networks, and SVM's. Linear/Logistic Regression and Naive Bayes are computationally very fast with a simple concept that make some parametric assumptions about the variable relationships, but work well as baseline models against which to compare complex models. In certain cases these simpler models are outperformed by complex models however it is important to compare both.

Ensemble tree methods are capable of performing as well as Neural Networks and SVM's and make no assumptions about the structure of the data, but provide insights into the model such as variable importance and variable relationships. Bayesian networks are excellent for visually presenting the structure and reasoning behind the predictive model, however are very slow to train and have been minimally included in the machine learning literature compared to the other methods. This may be due to low awareness of learned Bayesian Networks or lower performance. Neural Networks and SVM's are excellent machine learning predictors but are slow black box algorithms. Because it is so important to engage decision makers with a model that can explain its results, these two methods are not appropriate for the effort estimation problem.

None of the methods output statistical uncertainty or a confidence interval associated with each prediction. Understanding uncertainty should always be sought in decision making be it from a confidence interval or sensitivity testing. One way to achieve this would be to build hundreds of models using the same method but based on different training divisions of the data. This would produce a range of results for each case, and a statistical confidence interval or classification probability can be calculated for each prediction. Another way for a decision maker to get a feel for uncertainty is to perform sensitivity testing on a prediction. This means the user would run a case through a model several times, each time altering one of the explanatory variable to test the degree of variation in the model predictions. Both strategies would give an idea of the confidence with which a model is outputting response values.

In conclusion, simple models such as Linear/Logistic Regression and Naive Bayes are good baseline models to implement, while ensemble tree methods and Bayesian Networks may handle bigger unstructured data better than the simpler models while providing insight to decision makers.

>>>>

Engaging in loss-making jobs for fixed fees is a major problem in consulting[,]
"businesses."
"This is especially" 
[particularly] 
"true" in the competitive construction industry. This thesis investigates whether machine learning techniques applied to a company's passively collected internal data could help a 
[company] avoid loss-making jobs or 
"facilitate improved commercial strategy." 
[tactfully enforce stricter contracts.] 
It was found that in a specific decision framework, a case study's profits could be improved 9% by avoiding approximately 4% of projects.
"THIS IS AWESOME!!!! WHERE CAN I GET ONE?"
Alternative decision frameworks are also 
"proposed and evaluated" 
[debated and proposed]. The internal data collected by the case study company described each of their projects in terms of employee positions, fraction of project time completed by each employee, client characteristics, project characteristics (timespan, hours, technical details) and invoicing history. Algorithmic methods such as Logistic Regression, Random Forests, Boosted Trees, Naïve Bayes, and Bayesian Networks were applied as well as blended combinations of these methods. A decision scenario 
[that] 
"which" rejected projects above a sequence of tested thresholds was run in order to find the optimal threshold for profit improvements. This process was repeated on hundreds of models built from different training subsets of the data in order to obtain a 95% statistical confidence interval of profit improvements. The blended Logistic Regression model outperformed other methods and produced a 95% confidence interval of 6.5 - 11.5% profit improvements. Further work is recommended to test user input estimation and user interface designs. The findings from this research 
"has the potential to" 
[could] assist managers in reducing losses by highlighting risky projects and 
"guiding" 
[facilitating specific changes] "improvements" to their fee structures.

* why is it important?
affects the porfitability of a company which impacts the careers, company morale, and even viability of a company.

* why should someone read your essay?

* what is your main claim?

* discuss variables and your approaches

@@

@@

As stated, the aim of this project is to use statistical and machine learning techniques to model the profitability of projects. Several techniques will be tested and this section reviews a range of predictive algorithms that have been applied to business and social data. These vary from simple methods such as Linear Regression to complex, deep learning Neural Networks. The review of methods applied to the construction and IT industry highlighted the use of Linear Regression, Neural Networks, SVM's and in one case Boosted Trees, however research on other business problems utilised a wider range of methods. These included Naive Bayes, Random Forests, and machine learned Bayesian Networks. The following section presents an explanation fo each method, their advantages and disadvantanges and an example of a successful application.

@@

Linear Regression is one of the simplest and most popular statistical prediction methods. An equation for a line is learned, which is defined by a constant and coefficients multiplied by each explanatory variable. This can be mathematically represented by the equation below:

@@

where

$$
\begin{aligned}
X_{j}\ \textrm{represents the}\ j\textrm{th input variable} \\
\beta_{j}\ \textrm{is the coefficient or parameter being solved for} \\
\beta_{0}\ \textrm{is the y intercept}
\end{aligned}
$$

The most popular method to find the coefficients is to minimise the residual sum of squares (ordinary least squared method). The residuals represent the Y distance of each point from the line (predicted values), and the residual sum of squares is the sum of the squared value of each point's residual. It can be represented by the equation below:

@@

The reason for this method's popularity is that it is simple, provides stable predictions, and can be adapted for categorical variables [@Seng2010]. The integration of categorical variables is called Analysis of Variance (ANOVA) Linear Regression, which compares group mean variances within the categorical variables.

If structured correctly, Linear Regression results can show the importance of the explanatory variables. The variables must be first standardised to one another (i.e. scaled to the same mean and standard deviation). Then, their coefficients can be directly compared [@Putler2012]. The magnitude of the coefficients (the slopes) indicate how *much* a variable changes with the response variable. However, if two or more predictor variables are correlated, the values of their coefficients are unstable. More specifically, this means that if the a model has two correlated variables, and is re-run with a slightly different set of data, the coefficients for the two variables may be quite different. This is because explanatory variables are assumed to be independent, and if variables are correlated, it breaks this assumption. On the other hand, correlated variables do not impact the predictive accuracy of the model [@Putler2012].

In linear regression, no assumptions are made about the distributions of the variables. Skewed explanatory variable distributions are acceptable but it is generally better to normalise them by, for example, taking the logarithm or square root of the variable. The reason for this is that the more extreme values from the skewed distribution may create high leverage points and influence the slope (variable coeffiecients) too strongly [@Chatterjee1986]. The distribution of the residuals on the other hand, should always be Gaussian with a mean of 0. This is because if modeled well, the regression equation should follow the shape of the data [@Hastie2009].

Unfortunately many real world world phenomena do not have linear relationships which can make it difficult to produce accurate results using this method [@Breiman2001b]. There are methods to model non-linear relationships such as including polynomial terms and interaction terms. For example, in the linear equation, as well as $X_{1}$, include the terms $X_{2} = X_1^2$ and $X_{3} = X_1^3$ to build a third degree polynomial. Interactions between explanatory variables can be modeled by including a term such as $X_5$ where $X_5 = X_4*X_1$. This versatility is advantageous, however it can be challenging to know which non-linear terms of which variables to include. Sometimes data cannot be closely modelled by any equation and this is where other non-linear non-parametric methods can outperform linear regression.

@@

The ordinary least squared regression method described in the previous section cannot be applied to binary classification problems, where the objective is to predict the probability of an event occurring. Probability is a continuous response variable, however predictions from a linear regression would not be bound by 0 and 1. Instead, predicted values could reach further along the fitted line, outputting probabilities greater than 1 and less than 0 which is difficult to interpret. Also, residuals from linear regression should be normally distributed, but this is not possible when the response variable consists of only two values [@Lowry2016].

Nevertheless, the theory behind Linear Regression can be adapted to the prediction of probabilities pertaining to a binary response variable (represented by 1 or 0). First, a sigmoid function is assumed as the functional form of the probabilties against an example variable X. This is because it is pragmatic and works well: sigmoid functions can be bound by 0 and 1, and a logit transformation of the probability results in a linear relationship between the log odds and X. Refer to the equation below where *p* represents probability:

@@

The aim is to find which category has a higher probability, so to simplify calcualtions, the denominator can be dropped. Because the denominator is the same for both equations, it does not affect the final ratio of the probability of 'yes' to the probability of 'no'. Also integrating the assumption of conditional independence, the equations can be written as follows:

@@

These equations are simple to compute given the data, and the class with the highest probability can then be chosen [@Provost2013].

The advantages of this method are that the conditional independence assumption enables very fast calculations and predictions. The method can perform well in real world tasks because the assumption of independence does not significantly damage predictions. If multiple variables are actually related, the variables still separately direct the prediction in the correct direction. The correlated variables will double or triple their emphasis on the predicted response variable, however this does not necessarily impact classification accuracy because the final step is to choose the class with the *highest* probability [@Provost2013]. This is fine for ranking but the output probabilities are not accurate statistical probabilities [@Caruana2006]. Another disadvantage is that the distribution of numeric variables must be assumed (often Gaussian) in order to calculate the liklihood terms, and the data may not fit into these distributions neatly [@Caruana2006].

Example Gaussian probability densities of two response classes can be visualised on a bivariate feature space. The two normal distrubtions represent the probability of each class given values of the two explanatory variables (x and y axes):

@@

One real world success story for this classifier hinged on its ability to be refined incrementally with each piece of new data, as opposed to re-calibrating the entire model. Naive Bayes was used in a complex spam detection system where new spam emails or toxic text themes could be quickly added to the filtering model. Although the given example is different to cost estimation, the Naive Bayes method generally provides a good benchmark to compare against more complex models that should outperform it [@Caruana2006].

@@

Decision trees are one of the simplest and most intuitive machine learning methods. There are several brands of basic decision tree algorithms including ID3, C4.5, CART, and CHAID with CART and C4.5 being the most popular [@Kabra2011]. Ravi Kumar and Ravi [@Kumar2007] recommended the CART algorithm as it is capable of solving both classification and regression problems whereas the remaining decision trees solve classification problems only. Single decision trees create intuitive rules that a decision maker can follow in real-life scenarios but also tend to overfit the data and provide low predictive accuracy [@Kabra2011][@Putler2012]. Ensemble trees were developed to solve this problem and several ensemble techniques are described in this section.

Decision trees have a different theoretical foundation to traditional prediction methods such as regression. They are created by a progression of binary splits of the dataset that attempt to group similar response variable values. Before a split is made, each explanatory variable and each value within the explanatory variables is tested for how 'purely' it partitions the response variable into two subsets. For example, the explanatory variable 'gender' might split a response variable 'hair length' of a class of students into two subsets of students with more similar hair lengths. This similarity is called 'purity' and can be measured using several formulas. An example for regression (a problem with a numeric response variable) is:

$Purity = SSN - (SSL + SSR)$

and

@@

where

$$
\begin{aligned}
SSN\ \textrm{is the sum of squared errors of a node (dataset before the split)} \\
y_i\ \textrm{is the}\ i \textrm{th response variable value} \\
\bar{y}\ \textrm{is the mean response variable value in the node} \\
SSL\ \textrm{is the sum of squared errors in the left dataset that resulted from the split} \\ 
SSR\ \textrm{is the sum of squared errors in the right dataset that resulted from the split} 
\end{aligned}
$$

Other purity formulas also exist for categorical response variables such as Gini purity and information gain [@Hastie2009]. After the first split is made on the dataset, each subset is split by another variable (or the same explanatory variable again) and so on resulting in final subsets of data with similar response variable values (high purity). If this process is visualised graphically, it is clear that non-linear relationships can be captured.

@@

Once the splitting rules have been learned, a new case is run through the tree from the top. It follows the path provided by which side of the binary split its explanatory variable values fall under and finds its way to an end node. The mean response variable value from the training data that ended up in that node is the predicted response value for the new case. Decision trees are easy to overfit on the training data. This means that if the splitting rules are continued until no more splits can be possibly made, when the test data is run through the tree, the predictions will be too closely aligned to the exact structure of the training data. Prediction rules need to generalise trends found in the training data. This can be managed in a single tree by using 'stopping rules' such as a minimum number of data in a final node or a minimum purity increase for a split to occur [@Perlich2003].

As stated previously, single decision trees also suffer from low predictive accuracy and instability, so to combat these problems, ensemble tree methods were pioneered in the 1990's [@Breiman1996]. Three examples of these methods are boostrap aggregating (bagging), Random Forests, and Boosted Trees. Bagging builds multiple trees from different training sample data sets (wich are sampled from the full data set *with* replacement) and then combines tree results [@Breiman1996]. The method of combination is generally averaging numeric predictions or taking the majority rule for categorical variables.

@@

Lastly, the boosted decision tree approach applies gradient descent theory to a series of decision trees. The trees are limited to a certain depth to maintain simplicity, and each tree models the residuals (or errors) of the preceding tree. By modeling the errors, misclassified cases are weighted higher than correctly classified cases and influence the structure of the latest tree more [@Hastie2009]. This increased weighting on errors is why the algorithm is called boosting.

The boosting idea can be visualised in a simple scenario consisting of a single explanatory variable (shown on the x-axis) and response variable (y-axis). Each tree is limited to one split, and the first tree is shown below, where the vertical line is the splitting point:

@@

The horizontal lines indicate the mean response variable value in the left and right subset after the split. It is the value the first tree would predict for data that falls into each subset. The next tree is built on the residuals of the preceding tree. 
![Second tree modeling residuals in short progression of boosted trees [@Ihler2012].](images/lit_review/boost_tree2_.png)

@@

Further advantages of ensemble tree methods are numerous. Similarly to Linear Regression, the predictor variables in trees do not need to be transformed as no assumptions need to be made about the data's statistical distributions [@Louppe2014][@Radenkovic2010]. However, it is often wise to transform skewed distributions to achieve more even segmentation. Additionally, ensemble tree methods are faster than SVM's and Neural Networks but can perform just as well and even provide insights such as variable importance and variable relationships [@Sealfon2012]. Random Forests in particular have the benefit of being unaffected by noise in the data and have the capacity to handle a large number of variables [@Sealfon2012]. @Caruana2006 tested Boosted Trees, Random Forests, Neural Networks, SVM's, Logistic Regression and Naive Bayes on 11 binary classification problems and found that Boosted Trees performed best, followed by Random Forests. This demonstrates how ensemble tree methods are capable of competing with high level machine learning algorithms.

Although ensemble trees have numerous advantages it is just as important to understand their limitations. Trees are not built on a probabilistic framework, and therefore their results can not be provided in this framework. For example, statistical confidence intervals for predictions are not available for standard ensemble methods [@Louppe2014]. Solutions for this problem will be addressed in the summary of this section as the problem occurs in several methods. Also, variable importance tables can be output by Random Forests and Boosted Trees, but are often biased towards variables with many categories. Another type of ensemble method, conditional forests, should be used to check variable importance [@Radenkovic2010]. Finally, the methods can also be prone to overfitting if not tuned carefully [@Louppe2014][@Caruana2006].

@@

A Bayesian Network is a graphical probabilistic model that illustrates the conditional dependencies between variables in a data set. The model is visually represented by a directed acyclic graph (DAG) and is capable of linking the conditional dependency between any variable to another variable. arcs or edges represent links in the DAG [@Heckerman1998]. The conditional relationships are Bayesian, where the probabilities in one node are conditional upon values in nodes directed towards it as well as preceding nodes. Refer to the simple example below:

@@

If each variable, A, B, C, and D are binary categories, there are $2^4$, or 16 unique possible combinations of results and each variable is linked to each of the other variables with a conditionally dependency. The probability of each combination could be calculated using the chain rule for probabilities:

@@

However, domain knowledge or reasoning through the data could indicate that some variables are conditionally independent from one another. Then the calculations could be simplified [@Barber2012]. For example, if B and D, A and D, and A and C are conditionally independent, the DAG would be:

@@

In this way, Bayesian Networks reduce computations required to find the probability of a unique combination given other explanatory variable values while still taking into account many conditional dependencies that Naive Bayes, for example, cannot [@Barber2012]. Also, some equations for a comination are now clones of another combination and do not require recalculation. Calculating the conditional probability tables for each variable is still much more computationally expensive than Naive Bayes (as Naive Bayes is the simplest form of a Bayesian Netwoork) and variables with many categories dramatically increases the expense [@Zhang2004].

@@

Drawbacks of Bayesian Networks include the difficulty for them to process continuous variables. These demand sampling from conditional density models or complex algebra whereas categorical variables require comparitively simple calculations [@Hofmann1996]. It is therefore common to discretise continuous variables which simultaneously decreases precision of that variable [@Kragt2009].

Bayesian networks have found success in combining deterministic models with observed data as well as expert knowledge. Conditional probability dependencies can be calculated from different data sources and separately input into the network. They are therefore excellent at combining knowledge from different origins, and then visually communicating results to decision makers [@Kragt2009]. Another benefit of Bayesian Networks as decision support tools is they allow the user to fix certain variable values, re-run the network and observe changes in probabilities. This is a kind of sensitivity testing that gives users a feel for whether the model makes sense. Popular applications of Bayesian Networks are modeling uncertainty in natural resource management and modeling complex business network structures such as airports [@Wu2013].

@@

Algorithmic Neural Networks were inspired by Neural Networks of the human nervous system. In the human body, biological neurons recieve electro-chemical input into their dendrites from the terminals of adjacent neurons. If enough electro-chemical input is recieved, the neuron will 'fire', which means a signal will be sent from the dendrite end, through the axon to its terminal. Its terminal then sends signals to adjacent dendrites of other neurons and the process propagates. Neurons are able to learn over time by strengthening the connections of adjacent neurons that frequently cause the next one to fire. This means the required firing threshold for the recieving neuron lowers in strengthened connections [@Putler2012].

@@

A wide variety of neural network structures and dynamics exist, so a simple popular version is explained here. The first input layer is passive, and each input variable is passed through one input node. These nodes simply pass the input variable values to each of the nodes in the next layer, called a hidden layer. Neural networks may have any number of hidden layers with any number of nodes as selected by the analyst. Often however, a single hidden layer is adequate. Calibration for predictions is done in the hidden nodes, where each input variable is weighted then combined (similar to linear regression) to produce a single value. This is then transformed and assessed by a threshold function to determine if the hidden node should 'fire' to the next nodes. Once the signals and values have propogated to the final layer they are recombined to produce an output value, or a prediction [@Putler2012].

The algorithm learns from the data by starting with trial weights for each node. This produces a set of predictions from which errors are calculated. Then optimisation techniques are used to adjust the weights and reduce the error. This process can result in overfitting, so stopping rules such as regularisation strength should be used [@Karpathy2016], however these details are outside the scope of this review.

@@

SVM's are non-probabalistic learning algorithms that are typically used for binary classification, although multi-class and regression problems can be solved. The method works by geometrically determining the best decision boundary, called a hyperplane, between two classes in multi-dimensional space [@Byun2002] . A hyperplane is simply a plane in more than 3 dimensions, i.e. more than 3 explanatory variables. The simplest type of SVM resembles Logistic Regression because the hyperplane is a linear boundary, and like Logistic Regression, the category of future cases depends on which side of the boundary they fall.

@@

The plane is expressed with vectors because they can be defined such that $\boldsymbol{w}$ is perpendicular to $\boldsymbol{x}$. This aids in optimisation calculations to find the hyperplane with the widest margin, i.e. two hyperplanes with the largest perpendicular distance between them that contains less than the allowable number of misclassified points [@Fletcher2009]. The plot below demonstrates this concept and has no misclassifications:

@@

This section reviewed the advantages and disadvantages of statistical and machine learning techniques that have been successfully applied to business problems. These included Linear Regression, Logistic Regression, Naive Bayes, Decision Trees and Ensemble Trees, Bayesian Networks, Neural Networks, and SVM's. Linear/Logistic Regression and Naive Bayes are computationally very fast with a simple concept that make some parametric assumptions about the variable relationships, but work well as baseline models against which to compare complex models. In certain cases these simpler models are outperformed by complex models however it is important to compare both.

Ensemble Tree methods are capable of performing as well as Neural Networks and SVM's and make no assumptions about the structure of the data, but provide insights into the model such as variable importance and variable relationships. Bayesian Networks are excellent for visually presenting the structure and reasoning behind the predictive model, however are very slow to train and have been minimally included in the machine learning literature compared to the other methods. This may be due to low awareness of learned Bayesian Networks or lower performance. Neural Networks and SVM's are excellent machine learning predictors but are slow black box algorithms. Because it is so important to engage decision makers with a model that can explain its results, these two methods are not appropriate for the effort estimation problem.

None of the methods output statistical uncertainty or statistical confidence interval associated with each prediction. Understanding uncertainty should always be sought in decision making be it from a confidence interval or sensitivity testing. One way to achieve this would be to build hundreds of models using the same method but based on different training divisions of the data. This would produce a range of results for each case, and a statistical confidence interval or classification probability can be calculated for each prediction. Another way for a decision maker to get a feel for uncertainty is to perform sensitivity testing on a prediction. This means the user would run a case through a model several times, each time altering one of the explanatory variables to test the degree of variation in the model predictions. Both strategies would give an idea of the confidence with which a model is outputting response values.

In conclusion, simple models such as Linear/Logistic Regression and Naive Bayes are good baseline models to implement, while Ensemble Tree methods and Bayesian Networks may handle bigger unstructured data better than the simpler models while providing insight to decision makers.
